---
title: "Reading_and_Writing_data"
output:
  html_document:
    toc: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Subsetting Basics

There are a number of operators that can be used to extract subsets of R objects.  

**[** always returns an object of the same class as the original; can be used to select more than one
element (there is one exception)  

**[[ **is used to extract elements of a list or a data frame; it can only be used to extract a single
element and the class of the returned object will not necessarily be a list or data frame  
**$** is used to extract elements of a list or data frame by name; semantics are similar to that of [[.  

```{r }
x <- c("a", "b", "c", "c", "d", "a")
```


```{r }
x[1]
x[1:4]
 x[x > "a"]
  u <- x > "a"
  u
  x[u]
```

## Subsetting Lists

```{r}

x <- list(foo = 1:4, bar = 0.6)
x[1]
x[[1]]
x$bar
x[["bar"]]
x["bar"]


```
If extracting multiple elements, must use single brackets

```{r}

 x <- list(foo = 1:4, bar = 0.6, baz = "hello")
 x[c(1,3)]
 
```

The [[ operator can be used with computed indices; $ can only be used with literal names.

```{r}
x <- list(foo = 1:4, bar = 0.6, baz = "hello")
name <- "foo"
x[[name]] ## computed index for ‘foo’

x$name ## element ‘name’ doesn’t exist!

x$foo## element ‘foo’ does exist

```

### Subsetting Nested Elements of a List

The [[ can take an integer sequence.


```{r}
x <- list(a = list(10, 12, 14), b = c(3.14, 2.81))

x[[c(1, 3)]]

x[[1]][[3]]

x[[c(2,1)]]


```

##  Subsetting a Matrix

Matrices can be subsetted in the usual way with (i,j) type indices.

```{r}
x <- matrix(1:6, 2,3)
x[1,2]
x[2,1]


```

Indices can also be missing.

```{r}
x[1, ]
x[, 2]
```
By default, when a single element of a matrix is retrieved, it is returned as a vector of length 1 rather
than a 1 × 1 matrix. This behavior can be turned off by setting drop = FALSE


```{r}
 x <- matrix(1:6, 2, 3)
x[1,2]
x[1,2,drop=FALSE]


```
Similarly, subsetting a single column or a single row will give you a vector, not a matrix

```{r}
 x <- matrix(1:6, 2, 3)

x[1, ]
x[1, ,drop=FALSE]
```

## Removing NA VAlues
```{r}
 x <- c(1, 2, NA, 4, NA, 5)
bad <- is.na(x)
x[!bad]


```

What if there are multiple things and you want to take the subset with no missing values?

```{r}
x <- c(1, 2, NA, 4, NA, 5)
y <- c("a", "b", NA, "d", NA, "f")
good <- complete.cases(x, y)
good
x[good]
y[good]
```


## Vecorized Operations

Many operations in R are vectorized making code more efficient, concise, and easier to read.

```{r}
x <- 1:4; y <- 6:9
x + y
x > 2

x >= 2

y == 8

x * y

x / y

#------------------------
 x <- matrix(1:4, 2, 2); y <- matrix(rep(10, 4), 2, 2)
x
y

x*y #element wise multiplication

x/y

x%%y #True matrix multiplication




```





