---
title: "Functions"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## First Function

- write functions in a text file ( R script)


```{r}
add2 <- function(x,y){
  x+y
} ## didn't have to do anything to return, R returns the last expression
add2(3,5)
```

### Take a vector of numbers, return subset based on a condition

```{r}
above10 <- function(x){
  use <- x> 10 ##return logical vector
  x[use] ##return this
}
above <- function(x,n){
  use <- x > n
  x[use]
}

x<- 1:20
above(x,15)

above <- function(x, n = 10){ ## set default value to 10 
use <- x > n
x[use]
}


```

### Calculate the mean of each column in matrix or data frame


```{r}
columnmean <-function(y){ # y is a data fram or matrix
  nc <- ncol(y)
  means <-numeric(nc) #empty vector = num columns to store means
  for(i in 1:nc){
    means[i] <- mean(y[ , i])

  }
  means # return
}
d<- data.frame( col1=1:4, col2 = 5:8, col3 = c(NA, NA, 4, 6) )
d
columnmean(d)
#can't calculate mean where we have NA's in column

#rewrite the fn to remove NA's before calculting mean

columnmean <-function(y, removeNA = TRUE){ # y is a data fram or matrix
  nc <- ncol(y)
  means <-numeric(nc) #empty vector = num columns to store means
  for(i in 1:nc){
    means[i] <- mean(y[ , i], na.rm = removeNA)

  }
  means # return
}
columnmean(d)
columnmean(d, FALSE)



```


## Functions Part1

Functions are R objects of class 'function'  
The return value is the last expression in the function to be evaluated.  


Functions have **names argument** which may have default values.  
**Formal Arguments** : included in definition. 
**formals function** returns list of formal arguments.  

##Argument Matching


R functions arguments can be matched positionally or by name. So the following calls to sd are all
equivalent

```{r}
mydata <- rnorm(100)
sd(mydata)
sd(x=mydata)
sd(x=mydata, na.rm = FALSE) #when name your arguments, you don't have to use them in order
sd(na.rm = FALSE, x=mydata)
sd(na.rm = FALSE, mydata)
```


```{r}
args(lm)




#lm(data = mydata, formula=y-x, model=FALSE, subset=1:100)  
#lm(y-x, mydata, 1:100, model = FALSE) # if you don't overide defaulst, they will be used from the args defitinition 
```

## Functions Part 2

### Lazy evaluation

```{r}
f<- function(a,b){
  a^2
}
f(2) #fn never uses the value b, a is positionally matched to a
```


### The "..." Argument

The ... argument indicate a variable number of arguments that are usually passed on to other
functions.  
... is often used when extending another function and you don’t want to copy the entire argument
list of the original function 

```{r}
myplot <- function(x, y, type = "l", ...) {
 plot(x, y, type = type, ...)
}
```


The ... argument is also necessary when the number of arguments passed to the function cannot be
known in advance.  

```{r}
args(paste)
args(cat) # here the '...' is the set of R objects (we don't know how many ) that will be concatenated.
```

Any arguments that appear after ... on the argument list must be named
explicitly and cannot be partially matched.  


```{r}
args(paste)
paste("a", "b", sep =":")

```








