---
title: "Scoping Rules"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Symbol Binding

When you need to retrieve the value of an R object, the order is roughly:

```{r}
search()

```

The Global Environment = working environment


### Scoping Rules

Scoping rules determine how a value is associated with a free variable in a function

R Uses Lexical Scoping which means that values of free variables are searched for in the environment in which the function was defined.  
If not there, will go to parent environment ( next item on the search list)

Only when the value for the free variable does not exist in any environment, is an error thrown.  


## R Scoping Rules

Point: ** You can define things like global variables taht will be common to a lot of functions**

```{r}
make.power <- function(n){
  pow <- function(x){
    x^n
  }
  pow
}

#This function returns another function as its value
cube <- make.power(3)
cube(3)

# Whats in a  functions Environment?
ls(environment(cube))
get("n", environment(cube))



```
### Lexical vs. Dynamic Scoping

```{r}
y <- 10

f<- function(x){
  y<- 2
  y^2 +g(x)
}


g<-function(x){
  x*y # value of y in g function is 10.  This is lexical scoping.  Dynamic Scoping would assign y =2.
1
  }
```


Consequence of lexical scoping -> all R objects must be stored in memory



