---
title: "7. Matricies and data Framed"
output:
  html_document:
    df_print: paged
  word_document: default
---


 
 
 
 
## 7: Matrices and Data Frames
 
```{r}
my_vector <- 1:20
my_vector
dim(my_vector)
length(my_vector)


```
 The dim() function allows you to get OR set
| the `dim` attribute for an R object.
```{r}
dim(my_vector) <- c(4,5)
dim(my_vector)
attributes(my_vector)
my_vector


```
 my_vector is now a matrix
 
```{r}
class(my_vector)

```
 
 a matrix is simply an atomic vector with a
| dimension attribute. A more direct method of creating the
| same matrix uses the matrix() function.


```{r}
my_matrix <- my_vector
?matrix
my_matrix2 <- matrix(data=1:20, nrow=4, ncol=5)

```
 
The identical() function will tell us
| if its first two arguments are the same

```{r}
identical(my_matrix, my_matrix2)

```

Now, imagine that the numbers in our table represent some
| measurements from a clinical experiment, where each row
| represents one patient and each column represents one
| variable for which measurements were taken.
 We may want to label the rows, so that we know which
| numbers belong to each patient in the experiment. One way to do this is to **add a column to the matrix**, which contains| the names of all four people.

```{r}
patients <- c("Bill", "Gina", "Kelly", "Sean")
patients


```

use the cbind() function to 'combine columns'.

```{r}
cbind(patients, my_matrix)
```

Combining the character vector with our matrix of numbers
caused everything to be enclosed in double quotes. 
Matrices can only contain ONE class of data.

The data.frame() function allows us to store
 our character vector of names right alongside our matrix of numbers.
The data.frame() function takes any number of arguments and returns a single object of class `data.frame` that is composed of the original objects.



```{r}
my_data <- data.frame(patients, my_matrix)
my_data
class(my_data)

```

 Label Columns
 
```{r}
cnames <- c( "patient", "age", "weight", "bp", "rating","test")


```
 

 use the colnames() function to set the `colnames`
| attribute for our data frame

```{r}
colnames(my_data) <- cnames
my_data


```











