---
title: "subsetting vectors"
output:
  word_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Subsetting Vectors

 The way you tell R that you want to select some particular elements (i.e. a 'subset') from a vector
| is by placing an 'index vector' in square brackets immediately following the name of the vector.  
 
 Index vectors come in four different flavors -- logical vectors, vectors of positive integers,
| vectors of negative integers, and vectors of character strings -- each of which we'll cover in this.  

| lesson.

```{r}
x <- c( -2.14976472, -0.05860281, NA,NA, -0.30042357,  0.97797872, -1.57337945,  0.14209524, -0.42439809,          NA)
x
x[is.na(x)]
y <- x[!is.na(x)]
y

```

 Recall that the expression y > 0 will give us a vector of logical values the same length as y, with
| TRUEs corresponding to values of y that are greater than zero and FALSEs corresponding to values of y that are less than or equal to zero.

```{r}
y>0

y[y>0]

# Get positive elements of x that exclude NA
x[!is.na(x) & x >0]

#how we'd subset the 3rd, 5th, and 7th elements of x
x[c(3,5,7)]

#What if we're interested in all elements of x EXCEPT the 2nd and 10th?
x[c(-2,-10)]
x[-c(2,10)]




```

### 'named' elements.

```{r}
vect <- c(foo=11, bar=2, norf=NA)
vect

#We can also get the names of vect by passing vect as an argument to the names() function.
names(vect)

#Alternatively, we can create an unnamed vector vect2 with c(11, 2, NA)
#| Then, we can add the `names` attribute to vect2 after the fact with names(vect2) <- c("foo", "bar",| "norf")

vect2 <- c(11,2,NA)
names(vect2) <- c("foo", "bar", "norf")

#check that vect and vect2 are the same by passing them as arguments to the identical()
# function.
identical(vect,vect2)




```

### subsetting a vector by named elements

Get the second element of vect.

```{r}

vect["bar"]
#Likewise, we can specify a vector of names with vect[c("foo", "bar")]. Try it out.
vect[c("foo","bar")]


```





